\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-06 leslieandkenny.Rnw
Added some GLM vs GLMM commentary

require(xtable)
#require(foreign)
#require(arm)
require(tabplot)
require(car)
require(lme4)
require(lattice)
require(nlme)
@

<<data,echo=FALSE>>=
# Read and set up the data
fert2 <- read.csv("fert2.csv")

@@

\subsection{Binomial Generalized Linear Mixed Model}

\noindent Because the response variable is a Binomial count, and because
the design includes random assignment of variety to rows within blocks,
a binomial generalized linear mixed model is appropriate. This model
will estimate the probability of infection for each
\verb|variety:treatment:status| combination while controlling for
differences between blocks and between rows in the same block.

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
$$
\mathrm{logit}(p_i)=\mathbf{x}_i\mathbf{\beta}+b_{0,j[i]}+b_{1,j[i],k[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection,
\item \(\mathbf{x_i}\) is a row vector containing indicator variables for
the variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\mathbf{\beta}\) is a column vector of fixed-effect coefficients, and
\item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
\(b_{1,j,k}\sim\mathrm{N}(0,\sigma^2_{b_1})\) are random effects for block
and row, respectively.
\end{itemize}

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, and
that plots within a row are independent of each other.

\noindent Advantages of the Binomial GLMM include:
\begin{itemize}
\item It estimates and controls for variability between rows and blocks.
\item It avoids confounding variety effects with row effects and correctly
models the rows and nested within blocks.
\item Random effects can explain extra variation beyond what the
fixed-effect Binomial GLM can account for.
\end{itemize}

\noindent Disadvantages of this model include:
\begin{itemize}
\item It must be fit by numerically maximizing the likelihood with an
iterative algorithm. Including a large number of parameters (such as
third-order interaction coefficients) can cause the algorithm to fail
to converge.
\item The domain of the logit function is \((0, 1)\). If a large proportion
of the data are zeros, meaning an infection probability is near 0, the
algorithm will have difficulty estimating the coefficients. In this
situation, estimates may be unreliable or the algorithm may not converge.
\item We are not aware of any tools that can fit a Binomial GLMM with
overdispersion. If there are additional sources of variation that the
model does not account for, the standard error estimates will be too small.
\end{itemize}

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infrection rate. \textbf{If and
only if it is justified}, Nar could omit all of the data for this treatment
or variety when fitting the model. The scope of inference for this model
must then be limited to varieties and treatments that have a nonzero
infection rate.

\begin{figure}[b]
\caption{Example heatmap visualization for identifying patterns of zeros.}
<<heatmap,echo=FALSE,message=FALSE>>=
## Heatmap with meaningful ordering

# Order the dataset, using the arrange() function from the dplyr package
require(dplyr)
fert.ordered <- arrange(fert.sim1, variety, block, status, n.trt)

@@

\noindent If information about the physical layout of the field is
available, it would be possible to construct a similar heatmap to look
for spatial patterns. If correlation between nearby plots is seen,
the model should be modyfied to account for this.

@@

\begin{figure}[b]
\caption{Example \texttt{itableplot} output.}
<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
tableplot(fert.ordered, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = variety)
@
\end{figure}

@@

<<models1,warning=FALSE,results='hide',cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
                  family = binomial, data = fert.sim1)
@

We noticed that the SNMS variety had 27 out of 36 counts that were zeros.
Code like the following

<<models2,warning=FALSE,results='hide',cache=TRUE>>=
noVar1 <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
                family = binomial, data = fert.sim1, subset = variety!='Var1')
@

If there is no interest in comparing between inoculated plots and
control plots, separate models could be fit.

<<models3,warning=FALSE,results='hide',cache=TRUE>>=
onlyINOC <- glmer(cbind(infected, total) ~ variety*n.trt + (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
                  family = binomial, data = fert.sim1, subset = status=='INOC')

onlyCNTL <- glmer(cbind(infected, total) ~ variety*n.trt + (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
                  family = binomial, data = fert.sim1, subset = status=='CNTL')
@

<<models5,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
#summary(glmm3way)

glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
                    n.trt*status2 + (1|block/row),
                    control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
#summary(glmm2way)

#Anova(glmm3way)
#Anova(glmm2way)
#anova(glmm2way, glmm3way)
@

<<plots1,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert.sim1)
plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert.sim1)
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert.sim1$variety))
@

<<nosnms,echo=FALSE,warning=FALSE,results='hide',cache=FALSE>>=
table(fert2$infected==0,fert2$variety)

@@

<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

\begin{figure}
\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
require(effects)
plot(allEffects(glmm3way.sim), multiline = TRUE, type = 'response', se = TRUE,
     ci.style='bars', x.var = 'n.trt', rotx = 45, layout = c(5, 1),
     main = 'Variety:Treatment:Inoculation Effects Plot',
     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
     key.args = list(title = 'Inoculation Status'))
@
\end{figure}

<<glmmodel1,cache=FALSE>>=
glm1 <- glm(cbind(infected, total) ~ variety*n.trt*status2 + row,
            family = quasibinomial, data = fert.sim1)
glm2 <- glm(cbind(infected, total) ~ block + variety*n.trt*status2,
            family = quasibinomial, data = fert.sim1)
glm3 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
            family = quasibinomial, data = fert.sim1)
glm4 <- glm(cbind(infected, total) ~ block/row + variety*n.trt*status2,
            family = quasibinomial, data = fert.sim1)
glm5 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
            family = quasibinomial, data = fert.sim1)

glmnointMF <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
            family = quasibinomial, data = fert2)
glmmnointMF <- glmer(cbind(infected, total) ~ variety + variety*n.trt*status2 + (1|block/row),
            family = binomial, data = fert.sim1)
@

@@

>>>>

require(xtable)
#require(foreign)
#require(arm)
require(tabplot)
require(car)
require(lme4)
require(lattice)
@

<<data,echo=FALSE>>=
# Read and set up the data
fert2 <- read.csv('fert2.csv')

@@

\subsection{Binomial Generalized Linear Model}

\noindent Because the response variable is a Binomial count, and because
the design includes random assignment of variety to rows within blocks,
a Binomial generalized linear model is appropriate. We recommend using
a three-way interaction to estimate the probability of infection for each
variety/treatment/inoculation status combination and a block/variety
interaction to control for differences between blocks and between rows
in the same block.

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
$$
\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\).
\end{itemize}

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, and
that plots are independent given the other variables in the model.

\noindent There are several reasons why we recommend using a fixed-effects
generalized linear model instead of a generalized linear mixed model.
We observed overdispersion in Nar's data, and there are no simple tools
for including overdispersion in mixed models. Therefore, the standard
errors in the mixed model output would be based on an incorrect estimate
of the variance and might not be trustworthy. Additionally, including
many coefficients and nested effects in a mixed model can cause
computational problems where the software will fail to find unique
solutions for the parameters. These problems are made worse in a Binomial
GLMM when the counts are small, as in these data. The Binomial GLM can
adjust for overdispersion and avoid many of the computational difficulties
associated with the GLMM. The GLM can correctly model the nested
design through a block by variety interaction. The only advantage of the
GLMM is that it would estimate block- and row-level variances, but
it is our opinion that this is outweighed by the lack of an overdispersion
estimate and the computational difficulties.

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. A separate analysis would
be needed for more resistant population.

\begin{figure}
\caption{Example heatmap visualization for identifying patterns of zeros.}
<<heatmap,echo=FALSE,message=FALSE>>=
## Heatmap with meaningful ordering

# Order the dataset, using the arrange function from the dplyr package
require(dplyr)
fert.ordered <- arrange(fert.sim1, variety, block, status, n.trt)

@@

\noindent If information about the physical layout of the field is
available, it would be possible to construct a similar heatmap to look
for spatial patterns. If correlation between nearby plots is seen,
the model should be modified to account for this.

@@

\begin{figure}
\caption{Example \texttt{itableplot} output.}
<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = variety)
@
\end{figure}

@@

<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
@

If it is decided that certain observations come from an extremely
resistant population and can be omitted from the current analysis:

<<novar1,warning=FALSE,results='hide',cache=TRUE>>=
noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
              family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
@

If there is no interest in comparing between inoculated plots and
control plots, separate models could be fit:

<<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='INOC')

onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
@

<<models5,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
glm3way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
                family = quasibinomial, data = fert2)
summary(glm3way2)

glm2way2 <- glm(cbind(infected, total) ~ block*variety + (variety+n.trt+status2)^2,
                family = quasibinomial, data = fert2)
summary(glm2way2)

Anova(glm3way2, test.statistic = 'F')
Anova(glm2way2, test.statistic = 'F')
anova(glm2way2, glm3way2)
@

<<nosnms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=FALSE>>=
table(fert2$infected==0,fert2$variety)

@@

<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE>>=
par(mfrow = c(2, 2))
plot(resid(glm3way2) ~ predict(glm3way2), col = variety, data = fert2)
plot(resid(glm3way2) ~ fitted(glm3way2), col = variety, data = fert2)
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

\begin{figure}
\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
require(effects)
plot(Effect(c('variety','n.trt','status2'), glm3way),
     multiline = TRUE, type = 'response', se = TRUE, ci.style='bars',
     x.var = 'n.trt', rotx = 45, layout = c(5, 1),
     main = 'Variety/Treatment/Inoculation Effects Plot',
     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
     key.args = list(title = 'Inoculation Status'))
@
\end{figure}

@@

Adding a sentence to make sure I can push
